The benefits of reporting critical effect size values

### 2024-09-16 paper/paper.tex
updating how to compute es section

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The benefits of reporting critical effect size values},
  pdfauthor={Ambra Perugini1, Filippo Gambarota1, Enrico Toffalini2, Daniël Lakens3, Massimiliano Pastore1, Livio Finos4, Psicostat1, \& Gianmarco Altoè1},
  pdflang={en-EN},
  pdfkeywords={critical effect size values, statistical significance, fucntions implementation},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Where \(b\) is the unstandardized effect size that depends on the type of test. For example, in the two sample t-test or the single mean for a one sample t-test. The denominator is the standard error of the numerator.

@@

\subsubsection{two-sample t-test}\label{two-sample-t-test}

For the two sample t-test the critical value (\(b_c\)) is calculated using Equation \eqref{eq:eq-tw-sp}. Where \(t_c\) is the critical \(t\) value calculated using a certain \(\alpha\) value (e.g., \(0.05\)) and \(n_1 + n_2 - 2\) degrees of freedom. Then the \(b_c\) is simply divided by \(s\) obtaining the standardized critical effect size value. For the two-sample case (\(s = s_p\) assuming equal variances) Equation \eqref{eq:eq-s-p-twsmp} reported the pooled standard deviation.

\begin{equation}
    \label{eq:eq-tw-sp}
    b_c = t_c \times \text{SE}_{b}
\end{equation}

@@

When relaxing the assumption of equal variances (i.e., Welch's test) \(s\) is simply the square root of the average between the two variances.

\subsubsection{one sample t-test}\label{one-sample-t-test}

For the one sample case, the equations are the same. The only difference is that \(b\) is the sample mean and \(s\) is the sample standard deviation.

\subsubsection{paired sample t-test}\label{paired-sample-t-test}

For the paired sample case the situation is less straightforward. The reason is that the \(t\) statistic is computed using \(b\) being the average of the paired differences between the two (paired) samples and \(s\) being the standard deviation/error of the differences. While for hypothesis testing and power analysis it is common to calculate the \(d\) dividing by the standard deviation of the differences (Lakens, 2013), this effect size cannot be directly compared to the effect size used in the independent samples t-test, which divides by the pooled standard deviation (Morris \& DeShon, 2002). Practically, Equation \eqref{eq:eq-sp-sd} depicts the relationship between the pooled standard deviation and the standard deviation of the differences.

@@

Where \(\rho\) is the correlation between the two paired samples. Even if the hypothesis testing is computed using \(s_D\), we reported the critical effect size value both using the pooled standard deviation and the standard deviation of the differences. Beyond this difference, the equations are still the same as the previous example.

@@

The effect size calculated as in the previous step is known to be inflated thus we calculated also the corrected version usually known as Hedges's \(g\) (Hedges, 1981).

@@

For the Pearson's correlation test, the critical value is computed using Equation \eqref{eq:eq-cor-c} where \(t_c\) is still the critical value for the \(t\) distribution and \(n\) is the sample size.

\begin{equation}
    \label{eq:eq-cor-c}
    r_c = \frac{t_c}{\sqrt{n - 2 + t_c^2}}
\end{equation}

Another approach for hypothesis testing of the Pearson's correlation coefficient is using the Fisher's \(z\) transformation. Equation \eqref{eq:eq-f-rc} shows how to calculate the critical \(F(r_c)\) value where \(z_c\) is the critical value of the standard normal distribution with a chosen \(\alpha\) level. Then the \(F(r_c)\) can be transformed back into a correlation coefficient using Equation \eqref{eq:eq-bk-r}.

\begin{equation}
    \label{eq:eq-f-rc}
    \begin{gathered}
        F(r_c) = \frac{z_c}{\sqrt{n - 3}}
    \end{gathered}
\end{equation}

\begin{equation}
    \label{eq:eq-bk-r}
    r_c = \frac{\exp(2z_c)-1}{\exp(2z_c)+1} = \operatorname{tanh}(z_c)
\end{equation}

@@

Hypothesis testing on single coefficients in linear regression (using the \texttt{lm} function in R) is performed using the Wald test. This test is basically a t-test where the \(t\) value is calculated dividing the regression parameter \(\beta_0, \dots, \beta_j\) by the standard error. Then a \(t\) distribution using \(n - p - 1\) degrees of freedom where \(n\) is the number of observations and \(p\) is the number of coefficients beyond the intercept. The \(t\) value is calculated as reported in Equation \eqref{eq:eq-wald-lm} and the critical effect size value as reported in Equation \eqref{eq:eq-critical-lm}.

\begin{equation}
\label{eq:eq-wald-lm}
    t = \frac{\beta_j}{SE_{\beta_j}}
\end{equation}

\begin{equation}
\label{eq:eq-critical-lm}
    \beta_{jc} = t_c SE_{\beta_j}
\end{equation}

In this case \(\beta_{jc}\) is the unstandardized critical regression coefficient (as \(b\) in the \(t\)-test section). We implemented also the possibility to use standardized regression coefficients. We can identify two types of standardization, namely full and partial. The full standardization involves both the response variable and the predictors. Regression coefficients are interpreted as the increase in standard deviations of the response variable for an increase of one standard deviation in the predictor, keeping fixed all other predictors. The partial standardization involves only the predictors and regression coefficients are interpreted as the increase in the (raw) response variable for a increase in one standard deviation of the predictor. By default, standardizing means dividing a quantity by the standard deviation and eventually centering (i.e., subtracting the mean). Gelman (Gelman, 2008) suggested to standardize using two standard deviations because when numeric and binary variables are included the coefficients are not on the same metric. Equation \eqref{eq:eq-critical-lm} can be used but to calculate the critical value for the standardized regression coefficients the result need to be interpreted according to the chosen approach.

\subsubsection{Meta-analysis}\label{meta-analysis}

Meta-analysis allows to pool information from multiple studies related to a specific research question. The main advantage of meta-analysis is pooling multiple studies to obtain a more precise and powerful estimation of the effect. From a statistical point of view, a meta-analysis can be considered as a weighted linear regression with heterogeneous variances. Similarly to standard linear regression, hypothesis testing is performed using Wald \(t\) or \(z\) tests. The calculation of critical effect size values is the same as reported in Equations \eqref{eq:eq-wald-lm} and \eqref{eq:eq-critical-lm}. The only difference is that the standard error of the meta-analysis parameters included both the within-studies variances and eventually the between-studies heterogeneity.

\subsubsection{Other models}\label{other-models}

Despite we discussed only linear models, the same approach could be theoretically applied to other types of models such as generalized linear models. In fact, we simply need to multiply the critical effect size value of the chosen distribution (e.g., Student \(t\) or Standard Normal) by the standard error of the regression coefficient.

@@

@@

@@

@@

@@

@@

@@

>>>>

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother
\usepackage{xcolor}
\newcommand{\brev}{\color{red}}
\newcommand{\erev}{\color{black}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The benefits of reporting critical effect size values},
  pdfauthor={Ambra Perugini1, Filippo Gambarota1, Enrico Toffalini2, Daniël Lakens3, Massimiliano Pastore1, Livio Finos4, Psicostat1, \& Gianmarco Altoè1},
  pdflang={en-EN},
  pdfkeywords={critical effect size values, statistical significance, fucntions implementation},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Where \(b\) is the unstandardized effect size that depends on the type of test. For example, in the two sample t-test or the single mean for a one sample t-test. The denominator is the standard error of the numerator that depends on the sample size and the sample(s) variance.

@@

\color{red}

The critical \(t_c\) is the test statistic associated with a p-value equal or lower to \(\alpha\). Substituting \(t_c\) into Equation \eqref{eq:eq-gen-t} we realize that there is a \(b_c\) that with a certain \(SE_b\) produce the critical test statistics. The general form of the critical effect size \(d_c\) is defined in Equation \eqref{eq:eq-gen-crit}. The \(b_c\) is derived solving Equation \eqref{eq:eq-gen-t} for \(b\).

\begin{equation}
\begin{gathered}
    \label{eq:eq-gen-crit}
    d_c = \frac{b_c}{s} \\
    b_c = t_c \times SE_b
\end{gathered}
\end{equation}

NOT SURE TO KEEP THIS!

Another way of conceptualizing the critical effect size is by removing the sample size from the \(t\) statistics (see Equation \eqref{eq:eq-gen-t}). In fact, \(SE_b\) can be generally defined as \(\frac{s}{\sqrt{N}}\) where \(s\) is the denominator in Equation \eqref{eq:eq-gen-d} and \(n\) is the sample size. Thus, \(d_c = t_c \times \frac{1}{\sqrt{n}}\) (\(\frac{1}{\sqrt{n}}\) because the standard deviation of a standardized effect size is one by definition) is equivalent to \eqref{eq:eq-gen-crit}. The only caveat is that calculating \(d\) or \(d_c\) in this way assume that the standard deviation used to calculate the test statistics and the effect size is the same. While this is true most of the time, as explained in the paired t-test section, there are more than one methods to calculate the (critical) effect size using different standard deviations. Furthemore, this method assume that the two groups (in case of a two-sample t-test) have the same size.

\color{black}

\color{red}

\subsubsection{One sample t-test}\label{one-sample-t-test}

The easiest case is the one-sample t-test. \(b\) is the sample mean and \(s\) is the sample standard deviation and \(n\) is the sample size. \(SE_b = \frac{s}{\sqrt{n}}\) and the \(t_c\) is based on \(n - 1\) degrees of freedom. Thus \(b_c = t_c \times SE_b\) and \(d_c = \frac{b_c}{s}\). Equivalently, \(d_c = t_c \times \frac{1}{\sqrt{n}}\)

\color{black}

\subsubsection{Two-sample t-test}\label{two-sample-t-test}

We can apply the one-sample approach to the two-sample case. When assuming homogeneity of the variances between the two groups we have \(n_1 + n_2 - 2\) degrees of freedom to calculate \(t_c\) and \(s\) is the pooled standard deviation (\(s_p\)) calculated using Equation \eqref{eq:eq-s-p-twsmp}. The \(s_p\) can be intepreted as the square root of the weighted average of the samples standard deviations. The effect size \(d\) is calculated as \(\frac{b}{s_p}\) and the critical effect size can be calculated as \(d_c = \frac{b_c}{s_p}\) with \(b_c = t_c \times SE_b\). \(SE_b\) is calculated as \(s_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}\). As for the one-sample case, we can calculate the critical effect size from the t statistics as \(d_c = t_c \times \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}\).

@@

When relaxing the assumption of equal variances (i.e., using the Welch's t-test) \(s\) is calculated as the square root of the average between the two samples variances, \(SE_b\) is calculated as \(\sqrt{\frac{s^2_{x_1}}{n_1} + \frac{s^2_{x_2}}{n_2}}\). The degress of freedom for \(t_c\) are calculated in a more complex way using the Welch-Satterthwaite equation (Satterthwaite, 1946; Welch, 1947). For the Welch's t-test we can calculate the critical effect size removing the sample size from \(t_c\) (as done above) only when \(n_1 = n_2\).

\subsubsection{Paired sample t-test}\label{paired-sample-t-test}

For the paired sample case the situation is less straightforward. The reason is that the \(t\) statistic is computed using \(b\) being the average of the paired differences between the two (paired) samples and \(s\) being the standard deviation of the differences. Using these values, we can just compute the critical effects size as for the one-sample case, in fact a paired t-test is a one-sample test on the vector of paired differences. Test statistics and effect size based on the standard deviation of differences (the so called \(d_z\)) are used for hypothesis testing and power analysis (Lakens, 2013). The problem is that the \(d_z\) cannot be directly compared to the effect size calculated using the pooled standard deviation (Morris \& DeShon, 2002). If the correlation between the two paired samples is known there is a direct relationship between the pooled standard deviation \(s_p\) and the standard deviation of differences \(s_D\) as reported in Equation \eqref{eq:eq-sp-sd}

@@

Even if the hypothesis testing is computed using \(s_D\), we reported the critical effect size value both using the pooled standard deviation and the standard deviation of the differences.

@@

The effect size calculated as in the previous step is known to be inflated especially for small samples. For this reason, there is a corrected version of the effect size called Hedges's \(g\) Viechtbauer (2007). The Hedges's \(g\) can be calculated for all the t-test scenarios. The Hegdes's correction is implemented in Equation \eqref{eq:eq-g-correction} where \(\Gamma\) is the gamma function and \(m\) the degrees of freedom.

\begin{equation}
\label{eq:eq-g-correction}
c(m) = \frac{\Gamma\left(\frac{m}{2}\right)}{\sqrt{\frac{m}{2}} \Gamma\left(\frac{m-1}{2}\right)} \approx 1 - \frac{3}{4m - 1}
\end{equation}

The correction is applied as \(g = c(m) \times d\). The hypothesis testing is not taking into account the Hedges's correction thus the critical \(g\) will be different from the critical \(d\).

@@

Hypothesis testing for the Pearson's correlation coefficient is usually done using the t statistics or using a z statistics. The general approach presented for the t-test is still valid. The only difference is that the correlation is already a standardized effect size thus there is no need of the standardization term \(s\). Equation \eqref{eq:eq-gen-t} can be used substituing \(b\) with \(r\) (the sample correlation coefficient) as shown in Equation \eqref{eq:eq-rc}. The standard error is calculated as \(SE_r = \sqrt{\frac{1 - r^2}{n - 2}}\) (Bowley, 1928).

\begin{equation}
\label{eq:eq-rc}
    \begin{gathered}
        t = \frac{r}{SE_r} = r \sqrt{\frac{n - 2}{1 - r^2}} \\
        r_c = \frac{t_c}{\sqrt{n - 2 + t_c^2}}
    \end{gathered}
\end{equation}

Another approach for hypothesis testing of the Pearson's correlation coefficient is using the Fisher's \(z\) transformation (Fisher, 1915, 1921) reported in Equation \eqref{eq:eq-z-transf} and a \(z\) test statistics.

\begin{equation}
    \label{eq:eq-z-transf}
    \begin{gathered}
    r_z = \frac{1}{2} \ln \left(\frac{1 + r}{1 - r} \right) =                   \text{artanh}(r) \\
    r = \frac{\exp(2r_z - 1)}{\exp(2r_z + 1)} = \text{tanh}(r_z)
    \end{gathered}
\end{equation}

We can still use Equation \eqref{eq:eq-gen-t} substituting \(b\) with \(r_z\). The standard error of the \(r_z\) correlation is calculated as \(SE_{r_z} = \frac{1}{\sqrt{n - 3}}\). Equation \eqref{eq:eq-f-rc} shows how to calculate the critical \(r_{z_c}\) value where \(z_c\) is the critical value of the standard normal distribution with a certain \(\alpha\) level. Finally, \(r_{z_c}\) can be transformed back from the Fisher's \(z\) transformation using Equation \eqref{eq:eq-z-transf}

\begin{equation}
    \label{eq:eq-f-rc}
    \begin{gathered}
        z = \frac{r_z}{SE_z} = r_z \sqrt{n - 3} \\
        r_{z_c} = \frac{z_c}{\sqrt{n - 3}}
    \end{gathered}
\end{equation}

@@

Hypothesis testing on regression coefficients (e.g., using the \texttt{lm} function in R) is performed using the so-called Wald test. The test can be considered a one-sample t-test where the \(t\) value is calculated dividing the regression parameter \(\beta_0, \dots, \beta_j\) by its standard error \(SE_{\beta_j}\). The critical \(t\) value is calculated from a \(t\) distribution using \(n - p - 1\) degrees of freedom where \(n\) is the number of observations and \(p\) is the number of coefficients beyond the intercept. We can substitute \(b\) with \(\beta_j\) and \(SE_{b}\) with \(SE_{\beta_j}\) in Equation \eqref{eq:eq-gen-crit} to find the critical unstandardized regression coefficient (\(\beta_{c_j} = t_c \times SE_{\beta_j}\)). There is no an unique way to calculate the standardized version of the critical regression coefficient. The main reason is that in regression modelling we can include categorical and numerical independent variables with main effects and interactions. For example, Groß and Möller (2023) proposed a way to calculate a generalized Cohen's \(d\) from linear regression with or without the presence of other variables dividing the estimated parameter with the residual standard deviation. Gelman (2008) proposed to standardize using two standard deviations because when both numeric and categorical variables are included the coefficients are not on the same metric. Another complication is that in some cases, researchers standardize the dependent variable, the independent variables or both. In general, also the unstandardized critical regression coefficient can be usually easily intepreted. For example, assuming we are regressing some scores on a cognitive tests using the age of participant. The critical regression coefficient is the minimum increase in the cognitive test for a unit age increase that would be significant.

\subsection{Meta-analysis}\label{meta-analysis}

Meta-analysis allows to pool information from multiple studies related to a specific research question. The main advantage of meta-analysis is pooling multiple studies to obtain a more precise and powerful estimation of the effect. From a statistical point of view, a meta-analysis can be considered as a weighted linear regression with heterogeneous variances. Similarly to standard linear regression, hypothesis testing is performed using Wald \(t\) or \(z\) tests (Borenstein et al., 2009). For this reason, we can apply the same equations used for the linear regression. The main difference and advantage is that the meta-analysis model usually works directly with standardized effect size measures and regression coefficients are already standardized.

\subsection{Other models}\label{other-models}

Despite we discussed only linear models, the same approach could be applied to other types of models such as generalized linear models. In fact, we simply need to multiply the critical value of the chosen distribution (e.g., \(t\) or \(z\)) by the standard error of the regression coefficient. For example, in a logistic regression with a binary predictor the estimated regression coefficient is the (log) odds ratio. We can obtain the critical odds ratio by multiplying the critical \(z\) statistics (the default in R with the \texttt{glm} function) with the standard error of the regression coefficient.

@@

\footnotesize

@@

\normalsize

@@

\bibitem[\citeproctext]{ref-Borenstein2009-mo}
Borenstein, M., Hedges, L. V., Higgins, J. P. T., \& Rothstein, H. R. (2009). \emph{Introduction to {Meta-Analysis}}. \url{https://doi.org/10.1002/9780470743386}

\bibitem[\citeproctext]{ref-Bowley1928-yt}
Bowley, A. L. (1928). The standard deviation of the correlation coefficient. \emph{Journal of the American Statistical Association}, \emph{23}, 31. \url{https://doi.org/10.2307/2277400}

@@

\bibitem[\citeproctext]{ref-Fisher1915-ks}
Fisher, R. A. (1915). Frequency distribution of the values of the correlation coefficient in samples from an indefinitely large population. \emph{Biometrika}, \emph{10}, 507. \url{https://doi.org/10.2307/2331838}

\bibitem[\citeproctext]{ref-Fisher1921-qz}
Fisher, R. A. (1921). On the "probable error" of a coefficient of correlation deduced from a small sample. \emph{Metron}.

@@

\bibitem[\citeproctext]{ref-Gross2023-fl}
Groß, J., \& Möller, A. (2023). A note on cohen's d from a partitioned linear regression model. \emph{Journal of Statistical Theory and Practice}, \emph{17}. \url{https://doi.org/10.1007/s42519-023-00323-w}

@@

\bibitem[\citeproctext]{ref-Satterthwaite1946-aq}
Satterthwaite, F. E. (1946). An approximate distribution of estimates of variance components. \emph{Biometrics Bulletin}, \emph{2}, 110. \url{https://doi.org/10.2307/3002019}

@@

\bibitem[\citeproctext]{ref-Viechtbauer2007-tq}
Viechtbauer, W. (2007). Approximate confidence intervals for standardized effect sizes in the two-independent and two-dependent samples design. \emph{Journal of Educational and Behavioral Statistics: A Quarterly Publication Sponsored by the American Educational Research Association and the American Statistical Association}, \emph{32}, 39--60. \url{https://doi.org/10.3102/1076998606298034}

\bibitem[\citeproctext]{ref-Welch1947-wh}
Welch, B. L. (1947). The generalization of {``student's''} problem when several different population variances are involved. \emph{Biometrika}, \emph{34}, 28. \url{https://doi.org/10.2307/2332510}
